The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-14 paper/iphas-dr2.tex
Update iphas-dr2.tex

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy 
ensures that every pointing is immediately followed 
by an offset pointing at a displacement of $+$5 arcmin in Dec 
and $+$5 arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometric measurements
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were determined empirically by inspecting
the images and photometry by eye.
11\% of the data was rejected as part of this step.

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
%(e.g. bright satellite trails)
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1 arcsec.
The success of such a cross-matching procedure crucially depends
on the accuracy of the astrometric solution in each band,
and hence we were able to detect and correct
exposures with poor astrometry by inspecting the density map
for spurious density variations.

(6) \emph{Visual examination.}
All images and their associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey team members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images deemed unsuitable were flagged, investigated and
excluded from the release if necessary. 
6\% of the attempts to observe a field
were placed on a \emph{black-list}
for this reason, most commonly due to the obvious presence
of clouds or extreme levels of scattered moonlight.
Such fields were often rejected as part of 
one or more of the above quality criteria as well.

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field (H$\alpha$/$r$/$i$) 
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at nearly the same time --  
essentially always within 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap did not exceed 48 hours.
We note that the exact epoch of the magnitude in each band
is included in the source catalogue
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

The above criteria were satisfied 
for 14115 out of the 15270 fields (92\%).
In a few cases more than one successful attempt to observe
a field was available due to the fact that slightly stricter
quality criteria were adopted in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth was selected
for inclusion in the release, because the focus 
of this work is to deliver the most reliable
measurement at a single epoch.
Those interested in any of the rejected data 
may nevertheless access the full set of detection tables 
on line.

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relates to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr}
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(typically 0.09 for $r$ and 0.05 for $i$ in La Palma),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used.
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs 
and as such is an ancillary correction 
to account for sporadic gain variations. 
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

@@

Unlike the broadbands, 
standard star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for the narrowband.
We can specify here however 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter throughput curve 
corrected for atmosphere and detector quantum efficiency
\citep[using the method explained by][]{Drew2005}.
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric fluctuations during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic glitches
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[2MASS;][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[SDSS;][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
and Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 1--2\%.

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep{Padmanabhan2008,Schlafly2012}.

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
Firstly, the survey was carried out 
in competitive observing time
on a non-dedicated telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive
-- in part because standard fields 
are very scarce within the Galactic Plane.
Secondly, the aim of IPHAS is to obtain magnitudes at a single epoch
and hence stars are not normally observed at more than one epoch,
unless they happen to fall on the narrow overlap region 
between two neighbouring fields.

Although the IPHAS data does contain a significant number of 
inter-field repeat measurements
due to the small overlaps
between neighbouring fields,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is itself prone to systematics at the level of 1-2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is affected by a radial geometric distortion,
which causes the pixel scale towards the edges 
to increase \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine to cause artificial zeropoint gradients 
across the survey
when they are used to constrain a global calibration
without external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have instead opted to exploit an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.
This is explained next.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provided data across roughly half of the IPHAS footprint (Fig.~\ref{fig:apass_r}).
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS
for which we estimate the $1\sigma$-error to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of $\sim$3\%,
we used the APASS catalogue to identify and adjust all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.
For this purpose,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other hand.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the $r$ and $i$ filters 
are very similar in both surveys.
The transformations include a large fixed offset
which is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses magnitudes relative to Vega.
Separate transformations were derived for sightlines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes,
but the variations at these different
sightlines were found to be insignificant.
This is not surprising because heavily reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was the case for 48\% of the fields.
The mean offset was found to be
$0.014\pm0.104$ for $r$ and $0.007\pm0.108$ for $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

We then applied the most important step in our re-calibration scheme,
which is to adjust the provisional zeropoints of the 4596 aberrant fields
such that their offset is brought to zero.
This allowed the mean IPHAS-to-APASS offset 
to be brought down to $0.000\pm0.011$ in both $r$ and $i$
(Table~\ref{tbl:offsets_after}).
The procedure of fitting magnitude transformations and
correcting the zeropoints was repeated a few times to ensure 
convergence, which was essentially reached after the first iteration.

\begin{figure*}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_rshift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_r.pdf} 
    \caption{Median magnitude offset in the $r$ band between IPHAS and APASS,
             plotted on a field-by-field basis
             prior to the re-calibration procedure.
             Each square represents an IPHAS field
             which contains at least 30 stars with a counterpart
             in the APASS DR7 catalogue.
             The colours denote the median
             IPHAS-APASS magnitude offset in each field,
             which was computed after applying the APASS-to-IPHAS
             transformation to the APASS magnitudes (Eqn.~\ref{eqn:apass_r}).}
        \label{fig:apass_r}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_rshift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_r.pdf} 
    \caption{Median magnitude offset in the $r$ band
             between IPHAS and SDSS after the re-calibration
             procedure was applied.
             Again, each square represents an IPHAS field
             which contains at least 30 stars
             with a counterpart in the SDSS catalogue.
             The colours denote the median IPHAS-SDSS magnitude offset
             in each field,
             which was computed after applying the SDSS-to-IPHAS
             transformation to the SDSS magnitudes (Eqn.~\ref{eqn:sdss_r}).}
    \label{fig:sdss_r}
\end{figure*}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are pre-dominantly located 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme, explained below,
which minimises the magnitude offsets between stars
located in the overlap regions with neighbouring fields.
Although we explained earlier that these overlap regions
are prone to systematics, the use of APASS enabled us to
keep the zeropoints of roughly half of the fields fixed,
which avoids these systematics from combining to introduce
artificial gradients across the survey.
In the next section we will show this to be true by validating
our calibration against SDSS.

@@

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2
   \label{eqn:chi2}
\end{equation}
where $i$ denotes the exposure of interest, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$), 
$w_{ij}$ are weights set to the uncertainty in $\Delta_{ij}$
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$ which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}
This prescription is essentially identical to \citet{Glazebrook1994}.
%,although we changed the formulation to ensure that $A_{ij}$ is positive definite.

As explained above, we enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 6756 fields which have been compared
and calibrated against APASS earlier.
We will hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of these anchor fields 
are known and not solved for,
though they do appear in the vector $b_j$ as constraints.
In addition to the APASS-based anchors, 
we selected 3273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located at $\sim$500~meter from the INT.

We then solved Eqn.~\ref{eqn:matrix} for the $r$ and $i$ bands
separately using the least-squares routine 
in Python's {\sc scipy.sparse} module for sparse matrix algebra.
This provided us with corrected zeropoints for the remaining fields,
which were shifted on average by $+0.02\pm0.11$ in $r$ 
and $+0.01\pm0.12$ in $i$ compared to their provisional calibration.

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical
because the \ha\ zeropoints have been derived directly from the
$r$-band zeropoints during the provisional calibration (Eqn.~\ref{eqn:zpha}).
Moreover, the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken at essentially the same time, 
separated only by the $\sim$30-second read-out time required by the WFC.
We have hence corrected the \ha\ zeropoints 
by re-using the zeropoint adjustements that were derived for the $r$ band
in the earlier steps.
An exception was made for 3101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
which indicates non-photometric conditions.
For good practice, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.

>>>>

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy 
ensures that every pointing is immediately followed 
by an offset pointing at a displacement of $+$5 arcmin in Dec 
and $+$5 arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were determined empirically by inspecting
the images and photometry by eye.
We rejected 11\% of the data as part of this step
(many of which had already been rejected as part of the limiting magnitude criterion).

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
%(e.g. bright satellite trails)
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1 arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands.
In the majority of the cases we were able to correct the astrometry
of such fields by hand.

(6) \emph{Visual examination.}
All images and their associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images deemed unsuitable were flagged, investigated and
excluded from the release if necessary. 
6\% of the observations were placed on a \emph{black-list}
as part of this procedure, most commonly due to the obvious presence
of clouds or extreme levels of scattered moonlight.

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field (H$\alpha$/$r$/$i$) 
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at the $\sim$same time --  
essentially always within an elapsed time of 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap did not exceed 48 hours.
We note that the exact epoch of the magnitude in each band
is included in the source catalogue
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

The above criteria were satisfied by at least one observing attempt
for 14\,115 out of the 15\,270 fields (92\%).
In some cases more than one successful attempt to observe
a field was available due to the fact that stricter
quality criteria were adopted in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth was selected
for inclusion in the catalogue, because the focus 
of this release is to deliver the most reliable
measurement at a single epoch.
Those interested in any of the rejected data 
can nevertheless data-mine the full set of detection tables 
via our website.

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr}
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(typically 0.09 for $r$ and 0.05 for $i$ in La Palma),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used.
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs, 
and as such is an ancillary correction 
to account for sporadic gain variations. 
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

@@

Unlike the broadbands, 
standard star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for the narrowband.
We can specify here however 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter throughput curve 
corrected for the atmosphere and detector quantum efficiency
\citep[following the method explained by][]{Drew2005}.
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[2MASS;][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[SDSS;][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
and Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Padmanabhan2008,Schlafly2012}.

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
Firstly, the survey was carried out 
in competitive observing time
on a non-dedicated telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive
-- in part because standard fields 
are very scarce within the Galactic Plane.
Secondly, the aim of IPHAS is to obtain magnitudes at a single epoch
and hence stars are not normally observed at more than one epoch,
unless they happen to fall within a narrow overlap region 
between two neighbouring fields.

Although the IPHAS data does contain a significant number of 
such inter-field repeat measurements,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is most strongly affected by a radial geometric distortion in the WFC,
which causes the pixel scale towards the edges 
to increase \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine to cause artificial zeropoint gradients 
across the survey
when they are used to constrain a global calibration
without other external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have opted to involve an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.
This is explained next.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20\,cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides data across $\sim$half of the IPHAS footprint
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of $\sim$3\%,
we used the APASS catalogue to identify and adjust the calibration of all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.
For this purpose,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other hand.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the $r$ and $i$ filters 
are very similar in both surveys.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses magnitudes relative to Vega.
Separate transformations were derived for sightlines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sightlines were found to be insignificant.
This is not surprising because heavily reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was the case for 48\% of our fields.
The mean offset was found to be
$0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

We then applied the most important step in our calibration scheme,
which is to adjust the provisional zeropoints of these 4596 aberrant fields
such that their offset is brought to zero.
This allowed the mean IPHAS-to-APASS offset 
to be brought down to $0.000\pm0.011$~mag in both $r$ and $i$
(Table~\ref{tbl:offsets_after}).
The procedure of fitting magnitude transformations and
correcting the IPHAS zeropoints was repeated a few times to ensure 
convergence, which was essentially reached after the first iteration.

\begin{figure*}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_rshift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_r.pdf} 
    \caption{Median magnitude offset in the $r$ band between IPHAS and APASS,
             plotted on a field-by-field basis
             prior to the re-calibration procedure.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars with a counterpart
             in the APASS DR7 catalogue.
             The colours denote the median
             IPHAS-APASS magnitude offset in each field,
             which was computed after applying the APASS-to-IPHAS
             transformation to the APASS magnitudes (Eqn.~\ref{eqn:apass_r}).
             For clarity, we do not show the fields at the offset positions.}
        \label{fig:apass_r}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_rshift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_r.pdf} 
    \caption{Median magnitude offset in the $r$ band
             between IPHAS and SDSS after the re-calibration
             procedure was applied.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars
             with a counterpart in the SDSS catalogue.
             The colours denote the median IPHAS-SDSS magnitude offset
             in each field,
             which was computed after applying the SDSS-to-IPHAS
             transformation to the SDSS magnitudes (Eqn.~\ref{eqn:sdss_r}).
             For clarity, we do not show the fields at the offset positions.}
    \label{fig:sdss_r}
\end{figure*}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are pre-dominantly located 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme, explained below,
which minimised the magnitude offsets between stars
located in the overlap regions with neighbouring fields.
Although we explained earlier that these overlap regions
are prone to systematics, the use of APASS enabled us to
keep the zeropoints of roughly half of the fields fixed,
which avoids these systematics from combining to introduce
artificial gradients across the survey.
In the next section we will show this to be true by validating
our calibration against SDSS.

@@

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2
   \label{eqn:chi2}
\end{equation}
where $i$ denotes the exposure of interest, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the uncertainty in $\Delta_{ij}$
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$ which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}
This prescription is essentially identical to \citet{Glazebrook1994}.
%,although we changed the formulation to ensure that $A_{ij}$ is positive definite.

As explained above, we enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 6\,756 fields which have been compared
and calibrated against APASS earlier.
We hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of these anchor fields 
are known and not solved for,
though they do appear in the vector $b_j$ as constraints.
In addition to the APASS-based anchors, 
we selected 3\,273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located at $\sim$500\,meter from the INT.

We then solved Eqn.~\ref{eqn:matrix} for the $r$ and $i$ bands
separately using the least-squares routine 
in Python's {\sc scipy.sparse} module for sparse matrix algebra.
This provided us with corrected zeropoints for the remaining fields,
which were shifted on average by $+0.02\pm0.11$ in $r$ 
and $+0.01\pm0.12$ in $i$ compared to their provisional calibration.
The outcome is validated in the next section.

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical
because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken at essentially the same time, 
separated only by the $\sim$30\,sec read-out time required by the WFC.
We have hence corrected the \ha\ zeropoints 
by re-using the zeropoint adjustements that were derived for the $r$ band
in the earlier steps.
An exception was made for 3\,101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
suggesting strongly non-photometric conditions.
For good practice, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.
